---
title: "EVAR_machine_learning"
author: "Joshua Geiger and Mohammed Mehdi Shahid"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Update! Dataset is done, and ready to go!

Hey Mehdi,

Thanks for getting this started. The rds dataset that i provided is a
pretty good clean start. The variable that we will be attempting to
predict is sac_regression_12m_cat. This variable states weather The
Shins aortic aneurysms sac regressed states stable or increased in size
at the one-year mark. I had to do a bit of reformatting of all of the
data in order to calculate this appropriately given the variability in
the CT scan dates and sizes. I am confident this variable is accurate at
the moment.

There are variables within the data set that are incomplete and perhaps
chart reviewing to finish collecting some of this data will be
important. Particularly the first couple of patients were recorded
before all of the VQI variables were established and so there is some
missing data there. Additionaly, the last 20 some patients I recently
extracted from the VQI and so variables we manually extracted from the
chart have not be collected yet. Baqir and I can help out with this.
Baqir has also submitted a request for the lab values for these last 20
some patients. In addition we will be obtaining dates of death for the
entire data set. I think doing an additional analysis using similar
methods on survival, either length of survival or survival at 5 years
will also be interesting.

## SETUP

### Install Packages

```{r setup}
# Installs pacman ("package manager") if needed
if (!require("pacman")) install.packages("pacman")

#Install relevant libraries using pacman (i.e. package manager)

# Use pacman to load add-on packages as desired
pacman::p_load(pacman, 
               # essentials
               tidyverse, rio, caret,
               
               #plots
               ggplot2, mice, missRanger,
               
               #machine learning
               
               # extras
               forcats, janitor, explore, caTools, Metrics, VIM)
```

### Load data
```{r load-data}
#Sourcedir for Mehdi
EVARSourcedir <- "~/Desktop/EVAR/EVAR_cleaned_filtered_12_23_23.rds"

#Sourcedir for Josh
#EVARSourcedir <- "data/EVAR_cleaned_filtered_12_23_23.rds"

EVAR <- readRDS(EVARSourcedir)
```

### Correct Errors

```{r correct-errors}
#Correct errors in the EVAR dataframe
EVAR[ which(EVAR$primprocid == 798089), "height_cm"] <- 188.00
EVAR[ which(EVAR$primprocid == 798089), "bmi"] <- 32.7
```

### Pre-processing report
Generate a report about the data before any pre-processing
```{r}
EVAR_pre_report_filename <- "Report_EVAR_Preprocess.html"
EVAR_preprocess_report <- report(EVAR, 
                                 output_file = EVAR_pre_report_filename,
                                 output_dir = getwd())
```


## Cleaning Data

Remove cols and rows with too many missing values - remove records with
too many missing values (\>40%) - remove columns with too many missing
values (\>20%)

```{r}

# Remove records with too many missing values
EVAR_preprocess <- EVAR %>%
  remove_empty(which = "rows", cutoff = 0.6) #Remove observations with more than 60% missing data

# Remove columns with too many missing values
EVAR_preprocess <- EVAR %>%
  remove_empty(which = "cols", cutoff = 0.4) #Remove columns with more than 40% missing data

```

Remove unnecessary or uninformative columns

```{r}
EVAR_preprocess$zip_postal_code <- NULL
EVAR_preprocess$dysrhythmia <- NULL
EVAR_preprocess$primprocid <- NULL

#removing smoking quit date as I don't know how to work with this data type with ML for now....
#I created a years since quite smoking variable and removed the date. 
#EVAR_preprocess$quit_smoking_date <- NULL
```

### Data Types
Change column data types

When converting to factors make sure the ordering makes sense. For
example factor levels for smoking might look something like never, quit,
current. Changing the order to something like quit, current, never,
could very much mess with the analysis if these factors are considered
as ordinal variables, which they often are. can use fct_relevel to
change these

```{r}
# Convert all character columns to factors
EVAR_preprocess <- EVAR_preprocess %>%
  mutate_if(is.character, as.factor)

```
#### No and Yes variables
'No' and 'Yes' are mostly nominal, but for consistency, we can reorder
all factor variables with 'No' and 'Yes' to be No for level 1 and Yes
for level 2

```{r}
# Identify factor variables with 'No' and 'Yes' levels
no_yes_factors <- EVAR_preprocess %>%
  select(where(is.factor)) %>%
  keep(~ length(levels(.)) == 2 && all(c("No", "Yes") %in% levels(.)))

# Reorder the levels for 'No' as level 1 and 'Yes' as level 2
EVAR_preprocess <- EVAR_preprocess %>%
  mutate(across(all_of(names(no_yes_factors)), ~fct_relevel(., "No", "Yes")))

```

#### Releveling data
Determine which factors need to be releveled

```{r factor-variables}
# To make sure ordering of each factor variable makes sense, we must investigate each to determine which need to reordered with fct_relevel
# Extract information about factor variables
factorVars_info <- lapply(names(EVAR_preprocess), function(column_name) {
  column <- EVAR_preprocess[[column_name]]
  
  if (is.factor(column)) {
    levels_info <- levels(column)
    return(list(
      variable_name = column_name,
      num_levels = length(levels_info),
      levels = levels_info
    ))
  } else {
    return(NULL)  # Not a factor variable
  }
})

# Remove NULL elements (non-factor variables) from the list
factorVars_info <- factorVars_info[sapply(factorVars_info, function(x) !is.null(x))]

# Print the information
for (info in factorVars_info) {
  cat("Variable Name:", info$variable_name, "\n")
  cat("Number of Levels:", info$num_levels, "\n")
  cat("Levels:", paste(info$levels, collapse = "; "), "\n\n")
}
```

Based on the above output, we choose which variables to reorder and also
correct some errors within the dataset.

Correcting some errors
- aortic_graft_length_1

Convert to num
- aortic_graft_length_1
- right_iliac_graft_diameter_1
- right_iliac_graft_length_1
- left_iliac_graft_length_1

```{r correct-errors}
# aortic_graft_length_1 needs some error fixes and then convert to num
# Here we convert one of the errors from 90-30 to 90. Need to check if this is okay....

EVAR_preprocess <- EVAR_preprocess %>%
  mutate(
    # aortic_graft_length_1
    aortic_graft_length_1 = as.character(aortic_graft_length_1), 
    aortic_graft_length_1 = if_else(aortic_graft_length_1 == "90-30", 
                                    "90", aortic_graft_length_1),
    aortic_graft_length_1 = as.numeric(aortic_graft_length_1),
    
    
    # right_iliac_graft_diameter_1
    right_iliac_graft_diameter_1 = as.character(right_iliac_graft_diameter_1), 
    right_iliac_graft_diameter_1 = if_else(right_iliac_graft_diameter_1 == "16-14.5", 
                                           "16", right_iliac_graft_diameter_1),
    right_iliac_graft_diameter_1 = as.numeric(right_iliac_graft_diameter_1),
    
    
    #right_iliac_graft_length_1
    right_iliac_graft_length_1 = as.character(right_iliac_graft_length_1), 
    right_iliac_graft_length_1 = if_else(right_iliac_graft_length_1 == "Other", 
                                           NA, right_iliac_graft_length_1),
    right_iliac_graft_length_1 = as.numeric(right_iliac_graft_length_1),
    
    
    # left_iliac_graft_length_1
     left_iliac_graft_length_1 = as.character(left_iliac_graft_length_1), 
    left_iliac_graft_length_1 = if_else(left_iliac_graft_length_1 == "Other", 
                                           NA, left_iliac_graft_length_1),
    left_iliac_graft_length_1 = as.numeric(left_iliac_graft_length_1),
    )

```

Variables that need to be reordered: cvd, prior_chf, dysrhythmia,
diabetes, smoking, prior_pci, prior_cea_cas, prior_aneurysm_repair,
stress_test,pre_op_p2y12_antagonist, pre_op_beta_blocker,
pre_op_chronic_anticoagulant, ejection_fraction, aorta_neck_angle,
neck_aaa_angle, anesthesia, access_right, access_left,
right_iliac_adjunct, left_iliac_adjunct, proximal_aortic_extensions,
right_iliac_endpoint, left_iliac_endpoint, number_right,
distal_endpoint_right, number_left, distal_endpoint_left,
iv_bp_support_post_op, discharge_status

```{r reorder-factors}

#cvd
EVAR_preprocess$cvd <- fct_relevel(EVAR_preprocess$cvd,"None","Hx stroke, asymptomatic", "Hx stroke, minor deficit", "Hx stroke, major deficit")

#prior_chf
EVAR_preprocess$prior_chf <- fct_relevel(EVAR_preprocess$prior_chf,"None","Asymp, hx CHF", "Mild", "Moderate")

#copd
EVAR_preprocess$copd <- fct_relevel(EVAR_preprocess$copd,"No", "On home oxygen", "On meds", "Not treated")

#diabetes
EVAR_preprocess$diabetes <- fct_relevel(EVAR_preprocess$diabetes,"None","Diet","Non-insulin meds", "Insulin")

#smoking
EVAR_preprocess$smoking <- fct_relevel(EVAR_preprocess$smoking,"Never", "Prior", "Current")



#pre_op_p2y12_antagonist
EVAR_preprocess$pre_op_p2y12_antagonist <- fct_relevel(EVAR_preprocess$pre_op_p2y12_antagonist, "None", "No", "Clopidogrel","Ticagrelor", "Prasugrel" )

#pre_op_beta_blocker
EVAR_preprocess$pre_op_beta_blocker <- fct_relevel(EVAR_preprocess$pre_op_beta_blocker, "No", "No, for medical reason", "Pre-op 1-30 days", "Chronic > 30 days")

#pre_op_chronic_anticoagulant
EVAR_preprocess$pre_op_chronic_anticoagulant <- fct_relevel(EVAR_preprocess$pre_op_chronic_anticoagulant, "None", "No, for medical reason", "Other", "Rivaroxaban", "Warfarin", "Dabigatran")

#ejection_fraction
EVAR_preprocess$ejection_fraction <- fct_relevel(EVAR_preprocess$ejection_fraction,"Not done", "Unknown", "<30%", "30-50%", ">50%")


# Is aorta_neck_angle the same as neck_aaa_angle? 

#aorta_neck_angle
EVAR_preprocess$aorta_neck_angle <- fct_relevel(EVAR_preprocess$aorta_neck_angle, "< 45 Degrees", "45-60 Degrees", "61-75 Degrees", "76-90 Degrees", ">90 Degrees")
#neck_aaa_angle
EVAR_preprocess$neck_aaa_angle <- fct_relevel(EVAR_preprocess$neck_aaa_angle,"< 45 Degrees", "45-60 Degrees", "61-75 Degrees", "76-90 Degrees", ">90 Degrees")


#anesthesia
EVAR_preprocess$anesthesia <- fct_relevel(EVAR_preprocess$anesthesia,"Local", "Regional", "General")

#proximal_aortic_extensions
EVAR_preprocess$proximal_aortic_extensions <- fct_relevel(EVAR_preprocess$proximal_aortic_extensions, "None", "1", "2", "3")

#number_right
EVAR_preprocess$number_right <- fct_relevel(EVAR_preprocess$number_right,"None", "1", "2", "3")

```

## Data Preprocessing

### Split Data

```{r split-data}

set.seed(42)

# split using caTools package
split_sample <- sample.split(EVAR_preprocess$sac_regression_12m_cat, 
                             SplitRatio = 0.8)

# Subset train and test data frames
train_df <- subset(EVAR_preprocess, split_sample == TRUE)
test_df <- subset(EVAR_preprocess, split_sample == FALSE)
```

### Imputation

We assume data is missing at random (MAR).
The imputation strategy is to use MICE (Multivariate Imputation via Chained Equations) or missRanger.

#### Visualize missing data

```{r visualize-missing}
# Use VIM package
traindf_missing_plot <- aggr(train_df, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(train_df), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

testdf_missing_plot <- aggr(test_df, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(test_df), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))


```


Now We employ both and compare

```{r imputation}

train_imputed = list(
  imputed_mice <- train_df %>% mice(method = 'cart', m = 3, seed = 42) %>% complete(),
  imputed_ranger <- train_df %>% missRanger(formula = . ~ ., 
                                            pmm.k = 10, num.trees = 1000, 
                                            verbose = 2, seed = 42, 
                                            pred.var = c("sac_regression_12m_cat"))
  )

train_imputed = list(
  imputed_mice = test_df %>% mice(method = 'cart', m = 3, seed = 42) %>% complete(),
  imputed_ranger = train_df %>% missRanger(formula = . ~ ., 
                                            pmm.k = 10, num.trees = 1000, 
                                            verbose = 2, seed = 42, 
                                            pred.var = c("sac_regression_12m_cat"))
  )

```


```{r}
# Compare imputed datasets
# Example: Compare imputation accuracy using RMSE (Root Mean Squared Error)
rmse_mice <- rmse(train_df[, -1], complete_mice[, -1])  # calculate RMSE for MICE
rmse_missranger <- rmse(train_df[, -1], imputed_missranger$completed.data[, -5])  # calculate RMSE for missRanger

# Print results
cat("RMSE for MICE:", rmse_mice, "\n")
cat("RMSE for missRanger:", rmse_missranger, "\n")

```
