---
title: "EVAR_machine_learning"
author: "Joshua Geiger and Mohammed Mehdi Shahid"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
Update!
Dataset is done, and ready to go! there are 

Hey Mehdi,

Thanks for getting this started. The rds dataset that i provided is a pretty good clean start. The variable that we will be attempting to predict is sac_regression_12m_cat. This variable states weather The Shins aortic aneurysms sac regressed states stable or increased in size at the one-year mark. I had to do a bit of reformatting of all of the data in order to calculate this appropriately given the variability in the CT scan dates and sizes. I am confident this variable is accurate at the moment. 

There are variables within the data set that are incomplete and perhaps chart reviewing to finish collecting some of this data will be important. Particularly the first couple of patients were recorded before all of the VQI variables were established and so there is some missing data there. Additionaly, the last 20 some patients I recently extracted from the VQI and so variables we manually extracted from the chart have not be collected yet. Baqir and I can help out with this. Baqir has also submitted a request for the lab values for these last 20 some patients. In addition we will be obtaining dates of death for the entire data set. I think doing an additional analysis using similar methods on survival, either length of survival or survival at 5 years will also be interesting.


```{r setup}
# Installs pacman ("package manager") if needed
if (!require("pacman")) install.packages("pacman")

#Install relevant libraries using pacman (i.e. package manager)

# Use pacman to load add-on packages as desired
pacman::p_load(pacman, rio, dplyr, forcats, janitor, ggplot2, tidyverse, mice, caret)
```

```{r load-data}
EVARSourcedir <- "~/Desktop/EVAR/EVAR_cleaned_filtered_11_12_23.rds"
#Sourcedir for Josh
EVARSourcedir <- "data/EVAR_cleaned_filtered_12_23_23.rds"

EVAR <- readRDS(EVARSourcedir)
```


```{r correct-errors}
#Correct errors in the EVAR dataframe
EVAR[ which(EVAR$primprocid == 798089), "height_cm"] <- 188.00
EVAR[ which(EVAR$primprocid == 798089), "bmi"] <- 32.7
```


# DATA PREPROCESSING

Remove cols and rows with too many missing values
-   remove records with too many missing values (\>40%)
-   remove columns with too many missing values (\>20%)

```{r}

# Remove records with too many missing values
#EVAR_preprocess <- EVAR[complete.cases(EVAR) | rowSums(is.na(EVAR)) <= ncol(EVAR) * 0.4, ] # 0.4 denotes removing records with more than 40% missing values
#this can be shortened
EVAR_preprocess <- EVAR %>%
  remove_empty(which = "rows", cutoff = 0.6) #Remove observations with more than 60% missing data

# Remove columns with too many missing values
#EVAR_preprocess <- EVAR_preprocess[, colMeans(is.na(EVAR_preprocess)) <= nrow(EVAR_preprocess) * 0.2] # 20%
EVAR_preprocess <- EVAR %>%
  remove_empty(which = "cols", cutoff = 0.4) #Remove columns with more than 40% missing data

# Remove rows 430 to 461 as they are not complete yet
#Complete!
#EVAR_preprocess <- EVAR_preprocess[-(430:461), ]
```

Remove unnecessary or uninformative columns
```{r}
EVAR_preprocess$zip_postal_code <- NULL
EVAR_preprocess$dysrhythmia <- NULL

#removing smoking quit date as I don't know how to work with this data type with ML for now....
#I created a years since quite smoking variable and removed the date. 
#EVAR_preprocess$quit_smoking_date <- NULL
```

Change column data types

When converting to factors make sure the ordering makes sense. For example factor levels for smoking might look something like never, quit, current. Changing the order to something like quit, current, never, could very much mess with the analysis if these factors are considered as ordinal variables, which they often are. can use fct_relevel to change these

```{r}
# Convert all character columns to factors
EVAR_preprocess <- EVAR_preprocess %>%
  mutate_if(is.character, as.factor)

```

'No' and 'Yes' are mostly nominal, but for consistency, we can reorder
all factor variables with 'No' and 'Yes' to be No for level 1 and Yes
for level 2

```{r}
# Identify factor variables with 'No' and 'Yes' levels
no_yes_factors <- EVAR_preprocess %>%
  select(where(is.factor)) %>%
  keep(~ length(levels(.)) == 2 && all(c("No", "Yes") %in% levels(.)))

# Reorder the levels for 'No' as level 1 and 'Yes' as level 2
EVAR_preprocess <- EVAR_preprocess %>%
  mutate(across(all_of(names(no_yes_factors)), ~fct_relevel(., "No", "Yes")))

```

Determine which factors need to be releveled

```{r factor-variables}
# To make sure ordering of each factor variable makes sense, we must investigate each to determine which need to reordered with fct_relevel
# Extract information about factor variables
factorVars_info <- lapply(names(EVAR_preprocess), function(column_name) {
  column <- EVAR_preprocess[[column_name]]
  
  if (is.factor(column)) {
    levels_info <- levels(column)
    return(list(
      variable_name = column_name,
      num_levels = length(levels_info),
      levels = levels_info
    ))
  } else {
    return(NULL)  # Not a factor variable
  }
})

# Remove NULL elements (non-factor variables) from the list
factorVars_info <- factorVars_info[sapply(factorVars_info, function(x) !is.null(x))]

# Print the information
for (info in factorVars_info) {
  cat("Variable Name:", info$variable_name, "\n")
  cat("Number of Levels:", info$num_levels, "\n")
  cat("Levels:", paste(info$levels, collapse = "; "), "\n\n")
}
```

Based on the above output, we choose which variables to reorder and also correct some errors within the dataset.

Correcting some errors
aortic_graft_length_1 needs some error fixes and then convert to num
right_iliac_graft_diameter_1 convert to number
right_iliac_graft_length_1 and left_iliac_graft_length_1 - convert to num because has this 'Other' value


```{r correct-errors}
# aortic_graft_length_1 needs some error fixes and then convert to num
# Here we convert one of the errors from 90-30 to 90. Need to check if this is okay....
EVAR_preprocess$aortic_graft_length_1 <- as.character(EVAR_preprocess$aortic_graft_length_1)
EVAR_preprocess[which(EVAR_preprocess$aortic_graft_length_1 == "90-30"), "aortic_graft_length_1"] <- "90"
EVAR_preprocess$aortic_graft_length_1 <- as.numeric(EVAR_preprocess$aortic_graft_length_1)


# right_iliac_graft_diameter_1 to numeric
EVAR_preprocess$right_iliac_graft_diameter_1 <- as.double(EVAR_preprocess$right_iliac_graft_diameter_1)


# right_iliac_graft_length_1 convert to num and remove "other" option
# Convert "Other" to NA
for (convert in which(EVAR_preprocess$right_iliac_graft_length_1 == "Other")) {
  EVAR_preprocess[convert, "right_iliac_graft_length_1"] <- NA
}
EVAR_preprocess$right_iliac_graft_length_1 <- as.numeric(EVAR_preprocess$right_iliac_graft_length_1)

# left_iliac_graft_length_1 convert to num and remove "other" option
# Convert "Other" to NA
for (convert in which(EVAR_preprocess$left_iliac_graft_length_1 == "Other")) {
  EVAR_preprocess[convert, "left_iliac_graft_length_1"] <- NA
}
EVAR_preprocess$left_iliac_graft_length_1 <- as.numeric(EVAR_preprocess$left_iliac_graft_length_1)

```


Variables that need to be reordered: 
cvd, prior_chf, dysrhythmia, diabetes, smoking, prior_pci, prior_cea_cas, prior_aneurysm_repair, stress_test,pre_op_p2y12_antagonist, pre_op_beta_blocker, pre_op_chronic_anticoagulant, ejection_fraction, aorta_neck_angle, neck_aaa_angle, anesthesia, access_right, access_left, right_iliac_adjunct, left_iliac_adjunct, proximal_aortic_extensions, right_iliac_endpoint, left_iliac_endpoint, number_right, distal_endpoint_right, number_left, distal_endpoint_left, iv_bp_support_post_op, discharge_status


```{r reorder-factors}

#cvd
EVAR_preprocess$cvd <- fct_relevel(EVAR_preprocess$cvd,"None","Hx stroke, asymptomatic", "Hx stroke, minor deficit", "Hx stroke, major deficit")

#prior_chf
EVAR_preprocess$prior_chf <- fct_relevel(EVAR_preprocess$prior_chf,"None","Asymp, hx CHF", "Mild", "Moderate")

#diabetes
EVAR_preprocess$diabetes <- fct_relevel(EVAR_preprocess$diabetes,"No","Diet","Non-insulin med", "Insulin")

#dialysis
EVAR_preprocess$dialysis <- fct_relevel(EVAR_preprocess$dialysis,"No","Diet","Non-insulin med", "Insulin")

#smoking
EVAR_preprocess$smoking <- fct_relevel(EVAR_preprocess$smoking,"Never", "Prior", "Current")




#prior_pci
EVAR_preprocess$prior_pci <- fct_relevel(EVAR_preprocess$prior_pci,"Never", "Prior", "Current")

#prior_cea_cas
EVAR_preprocess$prior_cea_cas <- fct_relevel(EVAR_preprocess$prior_cea_cas,"Never", "Prior", "Current")

#prior_aneurysm_repair
EVAR_preprocess$prior_aneurysm_repair <- fct_relevel(EVAR_preprocess$prior_aneurysm_repair,"Never", "Prior", "Current")

#stress_test
EVAR_preprocess$stress_test <- fct_relevel(EVAR_preprocess$stress_test,"Never", "Prior", "Current")

#pre_op_p2y12_antagonist
EVAR_preprocess$pre_op_p2y12_antagonist <- fct_relevel(EVAR_preprocess$pre_op_p2y12_antagonist,"Never", "Prior", "Current")

#pre_op_beta_blocker
EVAR_preprocess$pre_op_beta_blocker <- fct_relevel(EVAR_preprocess$pre_op_beta_blocker,"Never", "Prior", "Current")

#pre_op_chronic_anticoagulant
EVAR_preprocess$pre_op_chronic_anticoagulant <- fct_relevel(EVAR_preprocess$pre_op_chronic_anticoagulant,"Never", "Prior", "Current")

#ejection_fraction
EVAR_preprocess$ejection_fraction <- fct_relevel(EVAR_preprocess$ejection_fraction,"Never", "Prior", "Current")

#aorta_neck_angle
EVAR_preprocess$aorta_neck_angle <- fct_relevel(EVAR_preprocess$aorta_neck_angle,"Never", "Prior", "Current")

#neck_aaa_angle
EVAR_preprocess$neck_aaa_angle <- fct_relevel(EVAR_preprocess$neck_aaa_angle,"Never", "Prior", "Current")

#anesthesia
EVAR_preprocess$anesthesia <- fct_relevel(EVAR_preprocess$anesthesia,"Never", "Prior", "Current")

#access_right
EVAR_preprocess$access_right <- fct_relevel(EVAR_preprocess$access_right,"Never", "Prior", "Current")

#access_left
EVAR_preprocess$access_left <- fct_relevel(EVAR_preprocess$access_left,"Never", "Prior", "Current")

#right_iliac_adjunct
EVAR_preprocess$right_iliac_adjunct <- fct_relevel(EVAR_preprocess$right_iliac_adjunct,"Never", "Prior", "Current")

#left_iliac_adjunct
EVAR_preprocess$left_iliac_adjunct <- fct_relevel(EVAR_preprocess$left_iliac_adjunct,"Never", "Prior", "Current")

#proximal_aortic_extensions
EVAR_preprocess$proximal_aortic_extensions <- fct_relevel(EVAR_preprocess$proximal_aortic_extensions,"Never", "Prior", "Current")

#right_iliac_endpoint
EVAR_preprocess$right_iliac_endpoint <- fct_relevel(EVAR_preprocess$right_iliac_endpoint,"Never", "Prior", "Current")

#left_iliac_endpoint
EVAR_preprocess$left_iliac_endpoint <- fct_relevel(EVAR_preprocess$left_iliac_endpoint,"Never", "Prior", "Current")

#number_right
EVAR_preprocess$number_right <- fct_relevel(EVAR_preprocess$number_right,"Never", "Prior", "Current")

#distal_endpoint_right
EVAR_preprocess$distal_endpoint_right <- fct_relevel(EVAR_preprocess$distal_endpoint_right,"Never", "Prior", "Current")

#number_left
EVAR_preprocess$number_left <- fct_relevel(EVAR_preprocess$number_left,"Never", "Prior", "Current")

#distal_endpoint_left
EVAR_preprocess$distal_endpoint_left <- fct_relevel(EVAR_preprocess$distal_endpoint_left,"Never", "Prior", "Current")

#iv_bp_support_post_op
EVAR_preprocess$iv_bp_support_post_op <- fct_relevel(EVAR_preprocess$iv_bp_support_post_op,"Never", "Prior", "Current")

#discharge_status
EVAR_preprocess$discharge_status <- fct_relevel(EVAR_preprocess$discharge_status,"Never", "Prior", "Current")

```





Hist visualization for numeric variables BEFORE imputation. Before we
impute missing data, we must visualize distributions of all numeric
variables and see if they change drastically with our imputation
strategy

Hist visualization for numeric variables BEFORE imputation.
Before we impute missing data, we must visualize distributions of all numeric variables and see if they change drastically with our imputation strategy

```{r}
# Extract column names (assuming the variables are numeric)
numeric_vars <- sapply(EVAR_preprocess, is.numeric)

# Set up a multi-plot layout
par(mfrow = c(2, 4), cex.main = 0.8, cex.lab = 0.6)  # Adjust the layout as needed

# Create histograms for all numeric variables pre imputation
preimputationHist <- for (var in names(EVAR_preprocess)[numeric_vars]) {
  hist(EVAR_preprocess[[var]], main = var, xlab = var, col = "lightblue", border = "black")
}

# Reset the layout to default
par(mfrow = c(1, 1),cex.main = 1, cex.lab = 1)
```

#### DATA SPLITTING

Note to future Mehdi: make sure you do NOT perform any feature selection
or rescaling on the entire dataset. These steps must happen for each
cross validation sequence so as to avoid data leakage. Here we shuffle
then split to create a hold out set. Each preprocessing step will be
done for both the train and test set to avoid data leakage

```{r}

set.seed(42)
train_percentage <- 0.8

# shuffle dataset
EVAR_preprocess <- EVAR_preprocess[sample(nrow(EVAR_preprocess)),]

indices <- createDataPartition(EVAR_preprocess$primprocid,
                               p = train_percentage,
                               list = FALSE)

# remove primprocid column as we no longer need it
EVAR_preprocess$primprocid <- NULL

trainSet <- EVAR_preprocess[indices, ]
testSet <- EVAR_preprocess[-indices, ]
```

#### DATA IMPUTATION

11. Numeric variables imputation.

The imputation strategy is to use MICE (Multivariate Imputation via Chained Equations) for numeric variables We assume data is missing at random (MAR)

```{r}

```
